Model Finding for (Co)recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-05 tex/main.tex

\begin{theorem}
If $\varphi$ be a $\Sigma$-formula that is in definitional form with respect to functions $\sfundefs{\Sigma}$,
and the set of function definitions corresponding to functions in $\sfundefs{\Sigma}$ is admissible,
then $\varphi$ and $\conv( \varphi )$ are equisatisfiable.
\end{theorem}
\begin{proof}
(Sketch)  First, we show that if $\varphi$ has a model $\I$, then $\conv( \varphi )$ has a model $\J$.
Let $\J$ be the $\extendsig{\Sigma}$-interpretation that interprets all types $\tau \in \ssorts{\Sigma}$ as $\tau^\I$,
all functions $\con{f} \in \sfuns{\Sigma}$ as $\con{f}^\I$,
and for each function $\con{f}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$, 
interprets $\fargsort{\con{f}}$ as a tuple of values of type $\tau_1^\I \times \ldots \times \tau_n^\I$,
and each $\fargx{\con{f}}{i}$ as the $i^{th}$ projection on such tuples for each $i = 1, \ldots, n$.
Note that $\J$ satisfies \emph{every} constraint of the form $\exists z\typ{\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( z ) \teq \vec t )$,
since by our construction of $\J$ there exists a $z$ such that $z^\J = \vec t^\J$.
By case analysis it follows that $\conv_0$ maintains the invariant that
if $\conv_0( t\typ{\Bool}, \ptrue )$ returns $( r\typ{\Bool}, D )$,
then $\I$ satisfies $t$ if and only if $\J$ satisfies $r$.
Thus, we conclude $\J$ is a model of $\conv( \varphi )$.

Second, we show that if $\conv( \varphi )$ has a model $\I$, then $\varphi$ has a model $\J$.

\end{proof}

>>>>

\begin{theorem}
If $\varphi$ is a $\Sigma$-formula that is in definitional form with respect to functions $\sfundefs{\Sigma}$,
and the set of function definitions $\mathcal{D}$ corresponding to functions in $\sfundefs{\Sigma}$ is admissible,
then $\varphi$ and $\conv( \varphi )$ are equisatisfiable.
\end{theorem}
\begin{proof}
(Sketch)  First, we show that if $\varphi$ is satisfied by $\Sigma$-interpretation $\I$, then $\conv( \varphi )$ is satisfied by a $\extendsig{\Sigma}$-interpretation $\J$.
Let $\J$ be the $\extendsig{\Sigma}$-interpretation that interprets all types $\tau \in \ssorts{\Sigma}$ as $\tau^\I$,
all functions $\con{f} \in \sfuns{\Sigma}$ as $\con{f}^\I$,
and for each function $\con{f}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$, 
interprets $\fargsort{\con{f}}$ as a tuple of values of type $\tau_1^\I \times \ldots \times \tau_n^\I$,
and each $\fargx{\con{f}}{i}$ as the $i^{th}$ projection on such tuples for each $i = 1, \ldots, n$.
Note that $\J$ satisfies \emph{every} constraint of the form $\exists z\typ{\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( z ) \teq \vec t )$,
since by our construction of $\J$ there exists a $z$ such that $z^\J = \vec t^\J$.
By case analysis it follows that $\conv_0$ maintains the invariant that
if $\conv_0( t\typ{\Bool}, \ptrue )$ returns $( r\typ{\Bool}, D )$,
then $\I$ satisfies $t$ if and only if $\J$ satisfies $r$.
Thus, we conclude $\J$ is a model of $\conv( \varphi )$.

Second, we show that if $\conv( \varphi )$ is satisfied by a $\extendsig{\Sigma}$-interpretation $\I$, then $\varphi$ is satisfied by $\Sigma$-interpretation $\J$.
Since $\varphi$ is in definitional form with respect to functions defined by $\mathcal{D}$, we know $\varphi$ is of the form $\mathcal{D} \wedge \varphi_0$.
We first argue that $\I$ satisfies a $\Sigma$-formula $\psi$ that is closed under function expansion with respect to $\mathcal{D}$,
and where $\psi$ entails $\varphi_0$.

Since there exists a model for $\psi$ that is closed under function expansion with respect to $\mathcal{D}$, 
then since $\mathcal{D}$ is admissible,
by definition there exists a model for $\psi \wedge \mathcal{D}$, which entails $\mathcal{D} \wedge \varphi_0$, which is $\varphi$.
\end{proof}

Let us revisit the translated formula~(\ref{eq:ex-after}) from Example~\ref{ex:translation}.
This formula is unsatisfiable if and only if formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargsort{f}$ can be interpreted as $\sortint$,
and $\farg{f}$ can be interpreted as the identity function.
Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if and only if formula~(\ref{eq:ex-before}) is satisfiable,
noting that the set $\{ \forall_{s} x\typ{\sortint}. ite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \}$ is admissble.
As stated in the proof of the previous theorem, 
the reason is that any model of formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under function expansion,
hence implying that formula~(\ref{eq:ex-before}) has a model.
In more detail, let $\varphi_s$ be the formula $\forall_{s} x. ite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) )$.
The rightmost conjunct 
$\exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c$ ensures 
that any model of formula~(\ref{eq:ex-after}) satisfying $s( c ) > 100$ also satisfies $\varphi_s[c/x]$.
Similarly, the nested existential $\exists z. \farg{s}( z ) \teq \farg{s}( y )-1 )$
ensures that this constraint is enforced recursively.
%Given a model $\I$ of formula~(\ref{eq:ex-after}) that interprets $c$ as $14$, it is easy to see that $\I$ also satisfies the formula:
\begin{eqnarray}
\begin{split}
s( c ) > 100 \wedge & \displaystyle\bigwedge\limits_{i=0}^{13} ( \neg c-i \leq 0 \wedge s( c-i ) \teq i + s( c-i-1 ) ) \\
                    & \wedge c-14 \leq 0 \wedge s( c-14 ) \teq 0
\end{split}
\end{eqnarray}
